---
title: "A Deep Dive into ANOVA (part 1)"
author: "Vadim Tyuryaev"
date: "2024-01-11"
output: html_document
---
  
```{r setup, include=FALSE}
# Set global chunk options
knitr::opts_chunk$set(echo = TRUE)
```

# Custom One-Way ANOVA Function

The following function, `one_way_anova`, takes a response variable and a grouping
variable, computes the necessary sums of squares, degrees of freedom, and the
F-statistic, and finally returns the p-value along with other ANOVA metrics.

```{r}

# Function to perform one-way ANOVA
one_way_anova <- function(response, group) {
  # Data preparation: create a data frame with the response and grouping variables.
  data <- data.frame(response, group)
  unique_groups <- unique(group)
  
  # Calculate the overall mean of the response variable.
  overall_mean <- mean(response)
  
  # Calculate sum of squares between groups (SSB)
  SSB <- sum(table(data$group) * (tapply(data$response, data$group, mean) - overall_mean)^2)
  
  # Calculate sum of squares within groups (SSW)
  SSW <- sum(tapply(data$response, data$group, function(x) sum((x - mean(x))^2)))
  
  # Degrees of freedom
  df_between <- length(unique_groups) - 1
  df_within <- length(response) - length(unique_groups)
  
  # Mean squares for between and within groups
  MS_between <- SSB / df_between
  MS_within <- SSW / df_within
  
  # Compute the F-statistic
  F <- MS_between / MS_within
  
  # Calculate the p-value from the F-distribution.
  p_value <- pf(F, df_between, df_within, lower.tail = FALSE)
  
  # Return the results as a list.
  return(list(F_statistic = F, 
              p_value = p_value, 
              df_between = df_between, 
              df_within = df_within))
}

```

# Analysis Using the [`mtcars`](https://stat.ethz.ch/R-manual/R-devel/library/datasets/html/mtcars.html) Dataset

We now load the built-in `mtcars` dataset and apply our custom one-way ANOVA
function to analyze the relationship between miles per gallon (MPG) and the
number of gears.

```{r}

# Load the mtcars dataset
data(mtcars)

# Apply the custom one-way ANOVA function on mpg and gear
result_custom_anova <- one_way_anova(mtcars$mpg, mtcars$gear)
print(result_custom_anova)

```

# Comparison with Built-In [`aov`](https://www.rdocumentation.org/packages/stats/versions/3.6.2/topics/aov) Function

We perform the same ANOVA using R's built-in `aov` function and compare the results.

```{r}

# Perform one-way ANOVA using the aov function
result_2 <- aov(mpg ~ factor(gear), data = mtcars)

# Display a summary of the ANOVA results
summary(result_2)

```

The results are **identical**. 

# BONUS: Groups Visualization with [`ggplot2`](https://ggplot2.tidyverse.org/)

We use the `ggplot2` package to create a boxplot that displays the distribution of
`MPG` across different gear levels. Different colors are assigned to each level,
and [jittered](https://ggplot2.tidyverse.org/reference/geom_jitter.html) data 
points are added for clarity.

```{r}

library(ggplot2)

# Create a boxplot with jittered points for each gear level
ggplot(mtcars, aes(x = factor(gear), y = mpg, fill = factor(gear))) +
  geom_boxplot() +
  geom_jitter(position = position_jitter(width = 0.2), alpha = 0.5) +
  # Add data points with slight horizontal jitter
  scale_fill_manual(values = c("red", "green", "blue")) +
  # Specify colors for each gear level
  labs(title = "Boxplot of MPG at Different Gears",
       x = "Gears",
       y = "MPG") +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5)  # Center the title
  )

```

What do you observe? 

